---
title: "Worksheet 05 Group 2"
author:
- "Andrea Staub"
- "Emanuel Mauch"
- "Holly Vuarnoz"
- "Jan Hohenheim"
- "Sophie Haldemann"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
# Set chunk options here 
knitr::opts_chunk$set(echo = TRUE)
```
# Exercise 3
Run the code provided in the ﬁle 05normal_example_JAGS.R and explore the interfaces in R to JAGS. Comment your ﬁndings. -> The comments can be found below the code chunks.



```{r}
###################
# Normal Example (Gibbs MCMC)
###################

remove(list=ls())
set.seed(44566)

# set the path to the 05normal_example_JAGS.txt file
path <- ""
```

clearing environment, `set.seed` for reproducibility and set path to .txt file with model

### Interface in R to JAGS: rjags

```{r}
##########################################################
# rjags interface to JAGS
##########################################################

library(rjags)

list.factories(type = "rng")
list.factories(type = "monitor")

list.factories(type = "sampler")
set.factory(name = "base::Slice", type = "sampler", state = FALSE)
list.factories(type = "sampler")
set.factory(name = "base::Slice", type = "sampler", state = TRUE)
list.factories(type = "sampler")

list.modules()
load.module("glm")
list.modules()
unload.module("glm")
list.modules()
```
With `list.factories` a data frame with two columns is returned, the first column shows the names of the factory objects in the currently loaded modules, and the second column is a logical vector indicating whether the corresponding factory is active or not.
With `list.modules()` the loaded modules can be listed. With `load.module("module_name")` and `unload.module("module_name")` modules can be loaded or unloaded.


```{r}
####################
## Introduction
####################


# generating data
#mu <- 4
#sigma2 <- 16
#n <- 30
#y <- rnorm(n=n, mean=mu, sd=sqrt(sigma2))


# Load the data
# round(y,3)
y <- c(3.048,2.980,2.029,7.249,-0.259,3.061,4.059,6.370,7.902,1.926,
       9.094,10.489,-0.384,-3.096,2.315,5.830,-1.542,-1.544,5.714,
       -5.182,3.828,-4.038,2.169,5.087,-0.201,4.880,3.302,3.859,
       11.144,5.564)

par(mfrow = c(1, 1))
boxplot(y)
summary(y)
sd(y)

# Define the parameters of the prior distributions 
mu0 <- -3
sigma2_0 <- 4
a0 <- 1.6
b0 <- 0.4
```

30 random datapoints were generated from a normal distribution with mean = 4 and standard deviation = 4. The parameters of the prior distribution are defined. Mean value from summary is with `r summary(y)[[4]]` not that close to mean = 4.

```{r}
################################
# INLA exact result (motivation)
################################

# https://www.r-inla.org/download-install
library(INLA)

formula <- y ~ 1
inla.output <- inla(formula,data=data.frame(y=y),
                    control.family = list(hyper =
                                            list(prec = list(prior="loggamma",param=c(a0,b0)))),
                    control.fixed = list(mean.intercept=mu0, prec.intercept=1/sigma2_0))


```

As input for INLA the previously defined parameters and the random sampled datapoints are used. INLA is used to approximate distribution.

```{r}
#############################
# Step 2: JAGS model file as a string in rjags with coda
#############################
set.seed(44566)

library(rjags)
library(coda)


#sessionInfo()

wb_data <- list( N=30, 
                 y=c(3.048,2.980,2.029,7.249,-0.259,3.061,4.059,6.370,7.902,1.926,
                     9.094,10.489,-0.384,-3.096,2.315,5.830,-1.542,-1.544,5.714,
                     -5.182,3.828,-4.038,2.169,5.087,-0.201,4.880,3.302,3.859,
                     11.144,5.564) 
)

wb_inits <- list( mu=-0.2381084, inv_sigma2=0.3993192 )

modelString = " # open quote for modelString
model{
# likelihood
for (i in 1:N){ 
y[i] ~ dnorm( mu, inv_sigma2 )    
}
# Priors
mu ~ dnorm( -3, 0.25 ) # prior for mu N(mu0, prec=1/sigma2_0)
inv_sigma2 ~ dgamma( 1.6, 0.4 ) # prior for precision G(a0, b0)

# transformations
# deterministic definition of variance
sigma2 <- 1/inv_sigma2

# deterministic definition of standard deviation
sigma <- sqrt(sigma2)
}
" # close quote for modelString

writeLines(modelString, con="TempModel.txt") # write to a file

```

The input for JAGS is defined as a list. The model is printed to a new .txt file. N is set to 30 and N samples are drawn from a normal distribution with mean mu and the precision as standard deviation. Mu follows a normal prior and the precision a gamma prior. The variance and then the standard deviation are estimated from the precision. 



```{r}
##########################################################
# JAGS only one chain
##########################################################

# model initiation
model.jags <- jags.model(
  file = "TempModel.txt", 
  data = wb_data,
  inits = wb_inits,
  n.chains = 1,
  n.adapt = 4000
)

str(model.jags)
class(model.jags)
attributes(model.jags)
list.samplers(model.jags)
```

A jags model with only one chain is initiated. The previously generated TempModel.txt file is used. The number of iterations for adaption is set to 4000.



```{r}
# burn-in

update(model.jags, n.iter = 4000)

# sampling
fit.jags.coda <- coda.samples(
  model = model.jags, 
  variable.names = c("mu", "sigma2", "inv_sigma2"), 
  n.iter = 10000,
  thin = 1
)

str(fit.jags.coda)
class(fit.jags.coda)
attributes(fit.jags.coda)

summary(fit.jags.coda)
#print(fit.jags.coda)
plot(fit.jags.coda)


# store samples for each parameter from the chain into separate objects
m.fit.jags.coda <- as.matrix(fit.jags.coda)
mu.sim <- m.fit.jags.coda[,"mu"] 
sigma2.sim <- m.fit.jags.coda[,"sigma2"]
inv_sigma2.sim <- m.fit.jags.coda[,"inv_sigma2"] 

library(MASS)

par(mfrow=c(1,1))
# plot for mean
rg <- range(inla.output$marginals.fixed$"(Intercept)"[,2])
truehist(mu.sim, prob=TRUE, col="yellow", xlab=expression(mu),ylim=rg)
lines(density(mu.sim),lty=3,lwd=3, col=2)
lines(inla.output$marginals.fixed$"(Intercept)",lwd=2)
legend("topright",c("MCMC: JAGS","INLA"),lty=c(3,1),lwd=c(2,2),col=c(2,1),cex=1.0,bty="n")

# plot for variance
m_var <-inla.tmarginal(function(x) 1/x, inla.output$marginals.hyperpar[[1]])
rg <- range(m_var[,2])
truehist(sigma2.sim, prob=TRUE, col="yellow", xlab=expression(sigma^2),ylim=rg)
lines(density(sigma2.sim),lty=3,lwd=3, col=2)
lines(m_var,lwd=2)
legend("topright",c("MCMC: JAGS","INLA"),lty=c(3,1),lwd=c(2,2),col=c(2,1),cex=1.0,bty="n")

# plot for precision
truehist(inv_sigma2.sim, prob=TRUE, col="yellow", xlab=expression(1/sigma^2))
lines(density(inv_sigma2.sim),lty=3,lwd=3, col=2)
lines(inla.output$marginals.hyperpar[[1]],lwd=2)
legend("topright",c("MCMC: JAGS","INLA"),lty=c(3,1),lwd=c(2,2),col=c(2,1),cex=1.0,bty="n")
```

The burn in period is set to 4000. After the burn in 10000 iterations are done to fit the model. The traceplots look good and stable. 
When looking at the traceplots we can see that the approximations for mu and sigma^2 from JAGS and INLA look similar but differ just a little bit. 



```{r}
##########################################################
# JAGS several chains
##########################################################

wb_inits <- function() {
  list(mu = rnorm(1),
       inv_sigma2 = runif(1)
  )  
}

# model initialisation
model.jags <- jags.model(
  file = "TempModel.txt", 
  data = wb_data,
  inits = wb_inits,
  n.chains = 4,
  n.adapt = 4000
)

# burn-in

update(model.jags, n.iter = 4000)

# sampling/monitoring
fit.jags.coda <- coda.samples(
  model = model.jags, 
  variable.names = c("mu", "sigma2", "inv_sigma2"), 
  n.iter = 10000,
  thin = 10
)

#n.thin<-floor((n.iter-n.adapt)/500)
#floor((10000-4000)/500)=12

```

Now again JAGS is used with the same model but instead of only one chain, 4 chains are used instead. The burn in period is again 4000.


```{r}
summary(fit.jags.coda)
plot(fit.jags.coda)



# store samples for each parameter from the chains into separate vectors
m.fit.jags.coda <-as.matrix(fit.jags.coda)
mu.sim <- m.fit.jags.coda[,"mu"] 
sigma2.sim <- m.fit.jags.coda[,"sigma2"]
inv_sigma2.sim <- m.fit.jags.coda[,"inv_sigma2"] 


par(mfrow=c(1,1))
# plot for mean
rg <- range(inla.output$marginals.fixed$"(Intercept)"[,2])
truehist(mu.sim, prob=TRUE, col="yellow", xlab=expression(mu),ylim=rg)
lines(density(mu.sim),lty=3,lwd=3, col=2)
lines(inla.output$marginals.fixed$"(Intercept)",lwd=2)
legend("topright",c("MCMC: JAGS","INLA"),lty=c(3,1),lwd=c(2,2),col=c(2,1),cex=1.0,bty="n")

# plot for variance
m_var <-inla.tmarginal(function(x) 1/x, inla.output$marginals.hyperpar[[1]])
rg <- range(m_var[,2])
truehist(sigma2.sim, prob=TRUE, col="yellow", xlab=expression(sigma^2),ylim=rg)
lines(density(sigma2.sim),lty=3,lwd=3, col=2)
lines(m_var,lwd=2)
legend("topright",c("MCMC: JAGS","INLA"),lty=c(3,1),lwd=c(2,2),col=c(2,1),cex=1.0,bty="n")

# plot for precision
truehist(inv_sigma2.sim, prob=TRUE, col="yellow", xlab=expression(1/sigma^2))
lines(density(inv_sigma2.sim),lty=3,lwd=3, col=2)
lines(inla.output$marginals.hyperpar[[1]],lwd=2)
legend("topright",c("MCMC: JAGS","INLA"),lty=c(3,1),lwd=c(2,2),col=c(2,1),cex=1.0,bty="n")


## CODA
#summary(fit.jags.coda)
#effectiveSize(fit.jags.coda)
#lapply(fit.jags.coda, effectiveSize)
#gelman.diag(fit.jags.coda,autoburnin=TRUE)
#gelman.plot(fit.jags.coda,autoburnin=TRUE)
#geweke.diag(fit.jags.coda)
#geweke.plot(fit.jags.coda)
#heidel.diag(fit.jags.coda)
#raftery.diag(fit.jags.coda)
#coda:::traceplot(fit.jags.coda)


# "DIC" penalised expected deviance computation
dic1<-dic.samples(model=model.jags, n.iter=1000, type="popt")
#dic2<-dic.samples(model=model.jags2, n.iter=1000, type="popt")

# "DIC" penalised expected deviance comparison
# There is no absolute scale for DIC comparison
# SE is very helpful

#diffdic(dic1,dic2)
```

The traceplots of the four chains are superimposed on each other. They look alright. Also this time the results from approximation with JAGS and INLA look quite similar.



### Interface in R to JAGS: runjags

```{r}

######
# Additional sampling in several chains, preparation for BGR/Gelman
# with runjags
######

library(runjags)

###############
# runjags interface with a link to a file
###############


wb_data <- list( N=30, 
                 y=c(3.048,2.980,2.029,7.249,-0.259,3.061,4.059,6.370,7.902,1.926,
                     9.094,10.489,-0.384,-3.096,2.315,5.830,-1.542,-1.544,5.714,
                     -5.182,3.828,-4.038,2.169,5.087,-0.201,4.880,3.302,3.859,
                     11.144,5.564) 
)

wb_inits <- function() {
  list(mu = rnorm(1),
       inv_sigma2 = runif(1)
  )  
}

fit.runjags<-run.jags(model=paste(path,"05normal_exmple_JAGS.txt",sep=""),
                      monitor=c("mu", "sigma2", "inv_sigma2"),
                      data=wb_data,
                      inits=wb_inits,
                      n.chains=4,
                      burnin=4000,
                      sample=5000,
                      adapt=1000,
                      thin=2)


plot(fit.runjags)
print(fit.runjags)

# CODA
fit.runjags.coda<-as.mcmc.list(fit.runjags)
summary(fit.runjags.coda)
# conduct CODA


```

Another interface is runjags. Now the .txt file 05normal_exmple_JAGS.txt is used where the model is specified. Four chains are used the burnin period is again set to 4000 afterwards 5000 additional samples are taken. In the burnin the adaptive iterations to use for the simulation are not included. Thinning is set to 2. Here burnin is directly specified in the model instead of using `update()`. The traceplots look ok. We also get an empirical cumulative distribution function plot and an autocorrelation plot for the parameters and an additional plot showing the correlations between the parameters. The autocorrelation plots look good for all parameters.



### Interface in R to JAGS: R2jags

```{r}
##########################################################
# R2jags wrapper to rjags interface to JAGS several chains
##########################################################



library(R2jags)

#rm(list=ls())

```

A third interface is R2jags.

```{r}
###############
# R2jags wrapper with a link to a file
###############


wb_data <- list( N=30, 
                 y=c(3.048,2.980,2.029,7.249,-0.259,3.061,4.059,6.370,7.902,1.926,
                     9.094,10.489,-0.384,-3.096,2.315,5.830,-1.542,-1.544,5.714,
                     -5.182,3.828,-4.038,2.169,5.087,-0.201,4.880,3.302,3.859,
                     11.144,5.564) 
)

#define parameters
params<-c("mu", "sigma2", "inv_sigma2")

# define inits
inits1 <- list(mu=rnorm(1), inv_sigma2=runif(1),
               .RNG.name="base::Super-Duper", .RNG.seed=1)
inits2 <- list(mu=rnorm(1), inv_sigma2=runif(1),
               .RNG.name="base::Wichmann-Hill", .RNG.seed=2)
wb_inits <- list(inits1,inits2)


fit.R2jags<-jags(data=wb_data,
                 inits=wb_inits,
                 parameters.to.save=params,
                 model.file=paste(path,"05normal_exmple_JAGS.txt",sep=""),
                 n.chains=2,
                 n.iter=50000,
                 n.burnin=4000,
                 n.thin=5,
                 DIC = TRUE,
                 jags.seed = 321,
                 refresh =100,
                 digits = 4,
                 jags.module = c("glm","dic"))


# Standard plots of the monitored variables
plot(fit.R2jags)

# Display summary statistics
print(fit.R2jags)

# traceplot
traceplot(fit.R2jags)


# CODA
fit.R2jags.coda<-as.mcmc(fit.R2jags)
summary(fit.R2jags.coda)
# conduct CODA
```

Now 2 Markov chains are used and a burnin of 4000. This time `n.iter` = 50000 includes the burnin period. The thinning is set to 5 to save memory and computation time. Also the model is called from the 05normal_exmple_JAGS.txt file. Again burnin is set directly instead of updating the model. Also this time the traceplots look alright.




# Exercise 4

Extend the code available in the file 05normal_example_JAGS.R to deal with the logistic
regression example for mice data from Collett (2003, p.71) provided in Table 1.

```{r}
###################
# Normal Example (Gibbs MCMC)
# Extention for logistic regression with JAGS
###################

remove(list=ls())
set.seed(44566)

# set the path to the 05normal_exmple_JAGS.txt file
path <- ""

##########################################################
# rjags interface to JAGS
##########################################################

library(rjags)

list.factories(type = "rng")
list.factories(type = "monitor")

list.factories(type = "sampler")
set.factory(name = "base::Slice", type = "sampler", state = FALSE)
list.factories(type = "sampler")
set.factory(name = "base::Slice", type = "sampler", state = TRUE)
list.factories(type = "sampler")

list.modules()
load.module("glm")
list.modules()
unload.module("glm")
list.modules()


####################
## Introduction
####################


# generating data
#mu <- 4
#sigma2 <- 16
#n <- 30
#y <- rnorm(n=n, mean=mu, sd=sqrt(sigma2))


# Load the data
# round(y,3)
y <- c(3.048,2.980,2.029,7.249,-0.259,3.061,4.059,6.370,7.902,1.926,
       9.094,10.489,-0.384,-3.096,2.315,5.830,-1.542,-1.544,5.714,
       -5.182,3.828,-4.038,2.169,5.087,-0.201,4.880,3.302,3.859,
       11.144,5.564)

# Mice data from Collett, 2003

# number of mice that survived
y2 <- c(26,9,21,9,6,1)

# drug concentration
x <- c(0.0028,0.0028,0.0056,0.0112,0.0225,0.0450)

# number of mice
n <- c(28,12,40,40,40,40)

par(mfrow = c(1, 1))
boxplot(y)
summary(y)
sd(y)

# Define the parameters of the prior distributions
# Normal example
mu0 <- -3
sigma2_0 <- 4
a0 <- 1.6
b0 <- 0.4

################################
# INLA exact result (motivation)
################################

# https://www.r-inla.org/download-install
library(INLA)

formula <- y ~ 1
inla.output <- inla(formula,data=data.frame(y=y),
                    control.family = list(hyper =
                                            list(prec = list(prior="loggamma",param=c(a0,b0)))),
                    control.fixed = list(mean.intercept=mu0, prec.intercept=1/sigma2_0))


#############################
# Step 2: JAGS model file as a string in rjags with coda
#############################
set.seed(44566)

library(rjags)
library(coda)


#sessionInfo()

wb_data <- list( N=30,
                 y=c(3.048,2.980,2.029,7.249,-0.259,3.061,4.059,6.370,7.902,1.926,
                     9.094,10.489,-0.384,-3.096,2.315,5.830,-1.542,-1.544,5.714,
                     -5.182,3.828,-4.038,2.169,5.087,-0.201,4.880,3.302,3.859,
                     11.144,5.564)
)

wb_inits <- list( mu=-0.2381084, inv_sigma2=0.3993192 )

modelString = " # open quote for modelString
model{
# likelihood
for (i in 1:N){
y[i] ~ dnorm( mu, inv_sigma2 )
}
# Priors
mu ~ dnorm( -3, 0.25 ) # prior for mu N(mu0, prec=1/sigma2_0)
inv_sigma2 ~ dgamma( 1.6, 0.4 ) # prior for precision G(a0, b0)

# transformations
# deterministic definition of variance
sigma2 <- 1/inv_sigma2

# deterministic definition of standard deviation
sigma <- sqrt(sigma2)
}
" # close quote for modelString

writeLines(modelString, con="TempModel.txt") # write to a file

##########################################################
# JAGS only one chain
##########################################################

# model initiation
model.jags <- jags.model(
  file = "TempModel.txt",
  data = wb_data,
  inits = wb_inits,
  n.chains = 1,
  n.adapt = 4000
)

str(model.jags)
class(model.jags)
attributes(model.jags)
list.samplers(model.jags)

# burn-in

update(model.jags, n.iter = 4000)

# sampling
fit.jags.coda <- coda.samples(
  model = model.jags,
  variable.names = c("mu", "sigma2", "inv_sigma2"),
  n.iter = 10000,
  thin = 1
)

str(fit.jags.coda)
class(fit.jags.coda)
attributes(fit.jags.coda)

summary(fit.jags.coda)
#print(fit.jags.coda)
plot(fit.jags.coda)


# store samples for each parameter from the chain into separate objects
m.fit.jags.coda <- as.matrix(fit.jags.coda)
mu.sim <- m.fit.jags.coda[,"mu"]
sigma2.sim <- m.fit.jags.coda[,"sigma2"]
inv_sigma2.sim <- m.fit.jags.coda[,"inv_sigma2"]

library(MASS)

par(mfrow=c(1,1))
# plot for mean
rg <- range(inla.output$marginals.fixed$"(Intercept)"[,2])
truehist(mu.sim, prob=TRUE, col="yellow", xlab=expression(mu),ylim=rg)
lines(density(mu.sim),lty=3,lwd=3, col=2)
lines(inla.output$marginals.fixed$"(Intercept)",lwd=2)
legend("topright",c("MCMC: JAGS","INLA"),lty=c(3,1),lwd=c(2,2),col=c(2,1),cex=1.0,bty="n")

# plot for variance
m_var <-inla.tmarginal(function(x) 1/x, inla.output$marginals.hyperpar[[1]])
rg <- range(m_var[,2])
truehist(sigma2.sim, prob=TRUE, col="yellow", xlab=expression(sigma^2),ylim=rg)
lines(density(sigma2.sim),lty=3,lwd=3, col=2)
lines(m_var,lwd=2)
legend("topright",c("MCMC: JAGS","INLA"),lty=c(3,1),lwd=c(2,2),col=c(2,1),cex=1.0,bty="n")

# plot for precision
truehist(inv_sigma2.sim, prob=TRUE, col="yellow", xlab=expression(1/sigma^2))
lines(density(inv_sigma2.sim),lty=3,lwd=3, col=2)
lines(inla.output$marginals.hyperpar[[1]],lwd=2)
legend("topright",c("MCMC: JAGS","INLA"),lty=c(3,1),lwd=c(2,2),col=c(2,1),cex=1.0,bty="n")














##########################################################
# JAGS several chains
##########################################################

wb_inits <- function() {
  list(mu = rnorm(1),
       inv_sigma2 = runif(1)
  )
}

# model initialisation
model.jags <- jags.model(
  file = "TempModel.txt",
  data = wb_data,
  inits = wb_inits,
  n.chains = 4,
  n.adapt = 4000
)

# burn-in

update(model.jags, n.iter = 4000)

# sampling/monitoring
fit.jags.coda <- coda.samples(
  model = model.jags,
  variable.names = c("mu", "sigma2", "inv_sigma2"),
  n.iter = 10000,
  thin = 10
)

#n.thin<-floor((n.iter-n.adapt)/500)
#floor((10000-4000)/500)=12

summary(fit.jags.coda)
plot(fit.jags.coda)



# store samples for each parameter from the chains into separate vectors
m.fit.jags.coda <-as.matrix(fit.jags.coda)
mu.sim <- m.fit.jags.coda[,"mu"]
sigma2.sim <- m.fit.jags.coda[,"sigma2"]
inv_sigma2.sim <- m.fit.jags.coda[,"inv_sigma2"]


par(mfrow=c(1,1))
# plot for mean
rg <- range(inla.output$marginals.fixed$"(Intercept)"[,2])
truehist(mu.sim, prob=TRUE, col="yellow", xlab=expression(mu),ylim=rg)
lines(density(mu.sim),lty=3,lwd=3, col=2)
lines(inla.output$marginals.fixed$"(Intercept)",lwd=2)
legend("topright",c("MCMC: JAGS","INLA"),lty=c(3,1),lwd=c(2,2),col=c(2,1),cex=1.0,bty="n")

# plot for variance
m_var <-inla.tmarginal(function(x) 1/x, inla.output$marginals.hyperpar[[1]])
rg <- range(m_var[,2])
truehist(sigma2.sim, prob=TRUE, col="yellow", xlab=expression(sigma^2),ylim=rg)
lines(density(sigma2.sim),lty=3,lwd=3, col=2)
lines(m_var,lwd=2)
legend("topright",c("MCMC: JAGS","INLA"),lty=c(3,1),lwd=c(2,2),col=c(2,1),cex=1.0,bty="n")

# plot for precision
truehist(inv_sigma2.sim, prob=TRUE, col="yellow", xlab=expression(1/sigma^2))
lines(density(inv_sigma2.sim),lty=3,lwd=3, col=2)
lines(inla.output$marginals.hyperpar[[1]],lwd=2)
legend("topright",c("MCMC: JAGS","INLA"),lty=c(3,1),lwd=c(2,2),col=c(2,1),cex=1.0,bty="n")


## CODA
summary(fit.jags.coda)
effectiveSize(fit.jags.coda)
lapply(fit.jags.coda, effectiveSize)
gelman.diag(fit.jags.coda,autoburnin=TRUE)
gelman.plot(fit.jags.coda,autoburnin=TRUE)
geweke.diag(fit.jags.coda)
geweke.plot(fit.jags.coda)
heidel.diag(fit.jags.coda)
raftery.diag(fit.jags.coda)
coda:::traceplot(fit.jags.coda)


# "DIC" penalised expected deviance computation
dic1<-dic.samples(model=model.jags, n.iter=1000, type="popt")
#dic2<-dic.samples(model=model.jags2, n.iter=1000, type="popt")

# "DIC" penalised expected deviance comparison
# There is no absolute scale for DIC comparison
# SE is very helpful

#diffdic(dic1,dic2)









######
# Additional sampling in several chains, preparation for BGR/Gelman
# with runjags
######

library(runjags)

###############
# runjags interface with a link to a file
###############


wb_data <- list( N=30,
                 y=c(3.048,2.980,2.029,7.249,-0.259,3.061,4.059,6.370,7.902,1.926,
                     9.094,10.489,-0.384,-3.096,2.315,5.830,-1.542,-1.544,5.714,
                     -5.182,3.828,-4.038,2.169,5.087,-0.201,4.880,3.302,3.859,
                     11.144,5.564)
)

wb_inits <- function() {
  list(mu = rnorm(1),
       inv_sigma2 = runif(1)
  )
}

fit.runjags<-run.jags(model=paste(path,"05normal_exmple_JAGS.txt",sep=""),
                      monitor=c("mu", "sigma2", "inv_sigma2"),
                      data=wb_data,
                      inits=wb_inits,
                      n.chains=4,
                      burnin=4000,
                      sample=5000,
                      adapt=1000,
                      thin=2)


plot(fit.runjags)
print(fit.runjags)

# CODA
fit.runjags.coda<-as.mcmc.list(fit.runjags)
summary(fit.runjags.coda)
# conduct CODA

















##########################################################
# R2jags wrapper to rjags interface to JAGS several chains
##########################################################



library(R2jags)

#rm(list=ls())


###############
# R2jags wrapper with a link to a file
###############


wb_data <- list( N=30,
                 y=c(3.048,2.980,2.029,7.249,-0.259,3.061,4.059,6.370,7.902,1.926,
                     9.094,10.489,-0.384,-3.096,2.315,5.830,-1.542,-1.544,5.714,
                     -5.182,3.828,-4.038,2.169,5.087,-0.201,4.880,3.302,3.859,
                     11.144,5.564)
)

#define parameters
params<-c("mu", "sigma2", "inv_sigma2")

# define inits
inits1 <- list(mu=rnorm(1), inv_sigma2=runif(1),
               .RNG.name="base::Super-Duper", .RNG.seed=1)
inits2 <- list(mu=rnorm(1), inv_sigma2=runif(1),
               .RNG.name="base::Wichmann-Hill", .RNG.seed=2)
wb_inits <- list(inits1,inits2)


fit.R2jags<-jags(data=wb_data,
                 inits=wb_inits,
                 parameters.to.save=params,
                 model.file=paste(path,"05normal_exmple_JAGS.txt",sep=""),
                 n.chains=2,
                 n.iter=50000,
                 n.burnin=4000,
                 n.thin=5,
                 DIC = TRUE,
                 jags.seed = 321,
                 refresh =100,
                 digits = 4,
                 jags.module = c("glm","dic"))


# Standard plots of the monitored variables
plot(fit.R2jags)

# Display summary statistics
print(fit.R2jags)

# traceplot
traceplot(fit.R2jags)


# CODA
fit.R2jags.coda<-as.mcmc(fit.R2jags)
summary(fit.R2jags.coda)
# conduct CODA
```

# Exercise 5

# Exercise 6

