---
title: "Presentation 5, Group 2: Classical Meta-Analysis of Binomial Outcomes"
author: "Jan Hohenheim"
date: "2024-05-14"
output:
  pdf_document: 
    latex_engine: xelatex
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
# Set chunk options here 

knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=F}
# Visual stuff
library(tidyverse)
library(broom)
library(ggthemes)
library(glue)
library(knitr)

# Meta analysis
library(lme4)
library(metafor)

theme_set(theme_solarized_2())
```

Source for theory: *Chang BH, Hoaglin DC. Meta-Analysis of Odds Ratios: Current Good Practices*

\newpage

## Data
Source: *Pagliaro L, D’Amico G, Sorensen TI, et al. Prevention of first bleeding in cirrhosis. A meta-analysis of randomized trials of nonsurgical treatment*

```{r, results='asis'}
dat.raw <- tibble(
  trial_id = 1:19,
  n_placebo = c(36, 53, 18, 22, 46, 60, 60, 69, 41, 20, 41, 35, 138, 51, 72, 16, 28, 19, 24),
  n_active = c(35, 56, 16, 23, 49, 53, 53, 71, 41, 21, 42, 33, 143, 55, 73, 13, 21, 18, 22),
  deaths_placebo = c(14, 29, 6, 6, 34, 14, 27, 26, 19, 2, 18, 21, 23, 24, 14, 4, 8, 6, 5),
  deaths_active = c(2, 12, 6, 4, 30, 13, 15, 16, 10, 0, 18, 20, 46, 19, 18, 2, 6, 7, 5)
)
dat.raw |>
  kable(caption = "Raw data")
```

```{r}
offset <- 0.001

dat <- dat.raw |>
  mutate(
    trial_id = as.factor(trial_id),
  ) |>
  mutate(
    p_placebo = (deaths_placebo + offset) / (n_placebo + 2 * offset),
    p_active = (deaths_active + offset) / (n_active + 2 * offset)
  ) |>
  mutate(
    log_odds_control = log(p_placebo / (1 - p_placebo)),
    log_odds_active = log(p_active / (1 - p_active))
  ) |>
  mutate(
    log_or = log_odds_active - log_odds_control,
    var_log_or =  
      1 / (deaths_placebo + offset) + 
      1 / (n_placebo - deaths_placebo + offset) + 
      1 / (deaths_active + offset) + 
      1 / (n_active - deaths_active + offset)
  )
```
```{r}
dat |> 
  filter(trial_id != 10) |>
  ggplot(aes(x = log_or, y = trial_id)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point(size = 3) +
  geom_errorbarh(aes(
    xmin = log_or - 1.96 * sqrt(var_log_or), 
    xmax = log_or + 1.96 * sqrt(var_log_or)), height = 0) +
  labs(
    title = "Forest plot of the included studies",
    x = "Log odds ratio",
    y = "Study ID"
  )
```

\newpage
## Methods

For binomial outcomes, the measured effects are usually transformed to the log-odds ratio.

$$
y_i := \log\left(\frac{p_i}{1-p_i}\right)
$$

Where $p_i$ is the probability of the event in group $i$.

The variance of the log-odds ratio is approximated by

$$
\text{Var}(y_i) \approx \frac{1}{a} + \frac{1}{b} + \frac{1}{c} + \frac{1}{d}
$$

Where $a, b, c, d$ are the number of events and non-events in the two groups:

```{r table2, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
|          | Responder | Non-Responder |
|----------|:---------:|:-------------:|
| Active   |   a       |   b           |
| Placebo  |   c       |   d           |
"
cat(tabl)
```

### Fixed effect
The measurement $y$ is an estimator of the underlying effect $\mu$ with an error term $\epsilon$. $\mu$ is assumed to be the same for all studies, while $\epsilon$ varies between studies.

$$
y_i = \mu + \epsilon_i, \quad \epsilon_i \sim \mathcal{N}(0, \sigma^2)
$$

The estimator $\hat{\mu}$ is the weighted average of the individual studies' estimates:

$$
\hat{\mu} = \frac{\sum \kappa_i y_i}{\sum \kappa_i}, \quad \kappa_i := \frac{1}{\sigma_i^2}
$$
Where $\sigma_i^2$ is the variance inside the study $i$.

If the number of subjects in each group is large, the variance of $\hat{\mu}$ can be approximated by

$$
\text{Var}(\hat{\mu}) \approx \frac{1}{\sum \kappa_i}
$$
In R, all of this is done quite simply with the `rma` function from the `metafor` package.

```{r, results='hold'}
rma.fixed <- rma(
  yi = dat$log_or, 
  vi = dat$var_log_or, 
  method = "FE") # FE stands for "Fixed Effect"

rma.fixed |> summary()
rma.fixed_mean <- rma.fixed$beta[[1]]
rma.fixed_se <- rma.fixed$se[1]
```
\newpage

### Random effects

$\mu$ can now also vary between studies:

$$
y_i = \mu_i + \epsilon_i, \quad \epsilon_i \sim \mathcal{N}(0, \sigma^2)
$$

Consequently, we now have to consider two sources of variation: The variance inside a study $\sigma_i^2$ and the variance between studies $\tau^2$:

$$
\text{Var}(y_i) \approx \sigma_i^2 + \tau^2
$$

The new estimator $\hat{\mu}'$ looks very similar to the fixed effect estimator, it only uses the new precision
$$
\hat{\mu}' = \frac{\sum \kappa_i' y_i}{\sum \kappa_i'}, \quad \kappa_i' := \frac{1}{\sigma_i^2 + \tau^2}
$$

Calculating the inter-study variance $\tau^2$ not as straightforward. `rma` lets you choose between different estimators, but the default is already fairly unbiased.

The mixed effects model is more flexible than the fixed effect model, but assumes that $y_i$ and $\sigma_i^2$ are independent, which is probably not the case for this kind of analysis.
As such, no matter which $\tau^2$ estimator is used, there will always be a bias. 

The R code looks very similar to the fixed effect model, with the difference being that we omit the `method` argument. The default already uses a random effects model.

```{r, results='hold'}
rma.random <- rma(
  yi = dat$log_or, 
  vi = dat$var_log_or)

rma.random |> summary()
rma.random_mean <- rma.random$beta[[1]]
rma.random_se <- rma.random$se[1]
```
\newpage

### Using sample sizes directly

If one has access to the sample size, the model can use the numbers of events and participants directly. In this case, one can do a simple logistic regression.
The response variable is the number of successes and events, and the explanatory variables are the trial number and whether the treatment was active or not.

```{r, results='asis'}
dat.glm <- tibble(
  trial_id = rep(dat$trial_id, 2),
  treatment = rep(c("placebo", "active"), each = nrow(dat)),
  deaths = c(dat$deaths_placebo, dat$deaths_active),
  n = c(dat$n_placebo, dat$n_active)
)
dat.glm |> 
  kable(caption = "Data for logistic regression")
```


```{r, results='hold'}
glm.fixed <- glm(
  cbind(deaths, n - deaths) ~ trial_id + treatment, 
  data = dat.glm, 
  family = binomial)

glm.fixed |> summary()

glm.fixed_placebo <- tidy(glm.fixed) |> 
  filter(term == "treatmentplacebo")
glm.fixed_mean <- -glm.fixed_placebo$estimate
glm.fixed_se <- glm.fixed_placebo$std.error
```
\newpage
Note that the above again assumes that $\mu$ is the same for all studies. We can extend this model to include a random effect for the study:


```{r, results='hold'}
numeric_treatment <-
  ifelse(dat.glm$treatment == "active", 1, 0)

glm.random <- glmer(
  cbind(deaths, n - deaths) ~ 
    trial_id + 
    treatment + 
    (-1 + numeric_treatment | as.numeric(trial_id)), 
  data = dat.glm, 
  family = binomial,
  nAGQ = 7)
glm.random |> summary(correlation=F)

glm.random_placebo <- summary(glm.random)$coefficients["treatmentplacebo",]
glm.random_mean <- -glm.random_placebo["Estimate"][[1]]
glm.random_se <- glm.random_placebo["Std. Error"][[1]]
```
\newpage
## Comparison

```{r}
comparison <- tibble(
  model = c("RMA Fixed effect", "RMA Random effects", "GLM Fixed effect", "GLM Random effects"),
  mean = c(rma.fixed_mean, rma.random_mean, glm.fixed_mean, glm.random_mean),
  se = c(rma.fixed_se, rma.random_se, glm.fixed_se, glm.random_se)
)

comparison |> kable(caption = "Comparison of the different models")
comparison |>
  ggplot(aes(x = model, y = mean, ymin = mean - 1.96 * se, ymax = mean + 1.96 * se)) +
  geom_pointrange() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_x_discrete(limits = c(
    "RMA Fixed effect", 
    "RMA Random effects", 
    "GLM Fixed effect", 
    "GLM Random effects")) +
  labs(
    title = "Comparison of the different methods",
    x = "Model",
    y = "Log odds ratio"
  ) +
  coord_flip()
```
