---
title: "Personal exercises session 1"
author: "Jan Hohenheim"
date: "2024-03-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, message = F}
library(tidyverse)
library(rjags)
library(coda)
library(bayesmeta)
library(pCalibrate)
```

# Exercise 1

## a)

The group investigates the safety and efficacy of Secukinumab for the treatment of ankylosing spondylitis compared to a placebo. 
Secukinumab is an antibody for IL-17, which is a pro-inflammatory cytokine believed to be involved in the pathogenesis of ankylosing spondylitis.

## b)

```{r}
dat <- matrix(data = c(14, 9, 1, 5), ncol = 2, byrow = T)
rownames(dat) <- c("Secukinumab", "Placebo")
colnames(dat) <- c("Response", "No response")
message("Data:")
print(dat)


message("Chi-squared test:")
chisq.test(dat)

message("Fisher's exact test:")
fisher.test(dat)
```

The reported p-values don't make me too excited about the results.

## c)

```{r}
power.prop.test(p1 = 0.6, p2 = 0.25, power = 0.8, sig.level = 0.05)
```

The reported optimal n = 30.1 matches neatly with the actual n = 30

# Exercise 2

My best guess, without cheating:  
- Mean: 172 cm  
- std-dev: 12 cm  

# Exercise 3

Everything works :)

# Exercise 4

## What is your motivation to attend this course?

I want to learn how to actually apply Bayesian methods to real data and how to interpret the results.
From talking to many people I've been convinced that this kind of statistics is highly relevant
to everyday life, as the human mind is a) very bad at understanding probabilities without guidance and b)
working in a Bayesian way by default. I've seen people apply Bayesian statistics as back-of-the-envelope calculations
for life decisions and I want to be able to do that too.

My MSc. is in neural systems and computation, where I have also encountered Bayesian statistics in the context of
neuroscience and machine learning. This left me largely scratching my head, as I didn't understand the jargon,
much less the content of the graphs and the meaning of the numbers involved. I want to change that.

## What would you like to learn in this course and why?

- How to do back-of-the-envelope calculations for life decisions.
- How to apply a Bayesian framework to biological data.

## Have you already learned or used Bayesian methods?

Only basic stuff, like the concept of a prior and a posterior, and the idea of a likelihood.
I've worked with Bayes' theorem in the context of probability theory and statistics, but it was not much to write home about.

## If yes, which Bayesian methods and programs are you familiar with?
N/A

## What is your experience with scientific writing?
I've written multiple scientific documents and incorporated feedback on all of them, so I think I'm alright.
